% RecSys 2025 — Long Paper (8 pages + 2 refs)
% Double-blind, double-column ACM sigconf format
\documentclass[sigconf,anonymous,review]{acmart}

\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath,amssymb,mathtools,amsthm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[capitalize,noabbrev]{cleveref}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}

\setcopyright{none}
\copyrightyear{2025}

\begin{document}

\title{Graph Positional Encoding and Variance-Weighted Learning\\as Universal Plug-ins for Uplift Modeling}

\begin{abstract}
Estimating Individual Treatment Effects (ITE) from networked observational data is critical for personalized recommendation, yet existing graph-based methods are monolithic—improvements to one architecture do not transfer to others. We introduce two orthogonal, plug-and-play modules that enhance \emph{any} graph ITE estimator: (1)~\textbf{Graph Positional Encoding (GPE)}, a multi-head self-attention fusion of node features and node2vec positional embeddings that captures community-level structural information, and (2)~\textbf{Variance-Weighted Learning}, an inverse-variance sample weighting scheme that mitigates heteroscedastic noise in treatment effect estimation. We evaluate both modules as drop-in enhancements on eight base models—including the foundational NetDeconf~(WSDM~2020) and the current state-of-the-art GDC~(WSDM~2025)—across five graph datasets and six real-world tabular uplift benchmarks, totaling over 1,100 experimental runs. GPE improves PEHE in 69\% of all graph settings, with TARNet+GPE achieving a perfect 11/11 win rate. On continuous outcomes, variance weighting yields a 4$\times$ Qini improvement. On high-noise graph settings, the combination of GPE and variance weighting reduces GDC's PEHE by up to 31\%. Our results establish that these modules constitute a general-purpose enhancement paradigm for graph-based uplift modeling in recommender systems.
\end{abstract}

\maketitle

%% ====================================================================
\section{Introduction}
%% ====================================================================

Accurate estimation of Individual Treatment Effects (ITE) is fundamental for personalized recommendations, from targeted advertising~\cite{parshakov2020spillover} to promotional campaigns~\cite{goldenberg2020free}. In networked environments such as social platforms, user behaviors are shaped not only by individual traits but also by peer influence through network connections~\cite{forastiere2021identification, aral2011creating}. This has motivated a growing body of work on graph-based ITE estimation, beginning with NetDeconf~\cite{guo2020learning} and progressing through GIAL~\cite{chu2021graph}, GNUM~\cite{zhu2023gnum}, and the current state-of-the-art GDC~\cite{hu2025gdc}.

Despite this progress, these methods share a fundamental limitation: each proposes a \emph{monolithic} architecture where improvements are tightly coupled to the specific model design. An enhancement that works for one architecture—say, a better loss function or a richer input representation—cannot be trivially transferred to another. This limits the practical utility of individual advances and forces practitioners to choose a single architecture rather than combining the best ideas.

We argue that the field would benefit more from \emph{modular, plug-and-play components} that improve any base model, rather than yet another end-to-end architecture. To this end, we propose two orthogonal modules:

\begin{itemize}
    \item \textbf{Graph Positional Encoding (GPE)}: A self-attention mechanism that fuses raw node features with node2vec~\cite{grover2016node2vec} positional embeddings. By making node positions explicitly available before GNN aggregation, GPE captures community-level structural biases that standard message-passing misses.
    \item \textbf{Variance-Weighted Learning}: An inverse-variance sample weighting scheme for the treatment effect regression loss. For continuous outcomes, the variance is learned via a dedicated head; for binary outcomes, we derive it analytically from the Bernoulli model ($\sigma^2 = \mu(1-\mu)$).
\end{itemize}

Both modules are \emph{architecture-agnostic}: GPE operates on the input (before the GNN backbone), while variance weighting operates on the loss (after the prediction heads). Neither modifies the base model's internal structure. We validate this universality through a rigorous ``control variable'' evaluation: we apply each module independently to eight different base models across eleven datasets, producing over 1,100 experimental runs.

Our key findings are:
\begin{enumerate}
    \item \textbf{GPE is a universal graph-uplift enhancer.} It improves PEHE in 69\% of all (model, dataset, noise) cells, including 7 of 8 tested base models. TARNet+GPE achieves a perfect 11/11 win rate across all graph datasets.
    \item \textbf{Variance weighting complements GPE under heteroscedasticity.} At high noise ($\rho=30$), GPE alone can overfit, but adding variance weighting stabilizes training. GDC+GPE+Var achieves the best PEHE on PubMed $\rho=30$ (2.99 vs.\ 4.31 vanilla).
    \item \textbf{On real-world tabular data}, variance weighting yields a 4$\times$ Qini coefficient improvement on continuous outcomes (Hillstrom spend).
\end{enumerate}

%% ====================================================================
\section{Related Work}
%% ====================================================================

\textbf{ITE estimation and uplift modeling.}
Meta-learners—S-learner, T-learner~\cite{shalit2017estimating}, X-learner~\cite{kunzel2019metalearners}, and DR-learner~\cite{kennedy2023towards}—form the backbone of modern ITE estimation. Deep learning extensions include BNN~\cite{johansson2016learning}, TARNet/CFRNet~\cite{shalit2017estimating}, SNet~\cite{curth2021nonparametric}, TNet~\cite{chen2024doubly}, and EFIN~\cite{liu2023explicit}. These methods assume i.i.d.\ data and do not account for network interference.

\textbf{Causal inference under network interference.}
Early work introduced partial interference~\cite{hudgens2008toward} and graph-cluster randomization~\cite{ugander2013graph}. NetDeconf~\cite{guo2020learning} pioneered GCN-based causal inference with representation balancing via Wasserstein distance. GIAL~\cite{chu2021graph} added mutual information maximization and adversarial balancing. GNUM~\cite{zhu2023gnum} proposed a transformed-target estimator for direct uplift prediction. IGL~\cite{sui2024invariant} introduced environment-invariant representations for inductive settings. GDC~\cite{hu2025gdc} achieves the current SOTA by disentangling node features into adjustment and confounder factors with treatment-conditioned graph aggregation.

\textbf{Heteroscedasticity in ITE estimation.}
Variance-weighted methods have been explored in tabular settings~\cite{SHI2024111679, mo2022efficient, abrevaya2023estimation} but have not been integrated with graph-based approaches. Our work bridges this gap.

\textbf{Our contribution.} Unlike prior work that proposes a new monolithic model, we introduce two modular enhancements and validate them across the full spectrum of existing models, from NetDeconf (2020) to GDC (2025).

%% ====================================================================
\section{Preliminaries}
%% ====================================================================

\subsection{Uplift Modeling}
For each user $i$ with covariates $X_i \in \mathbb{R}^d$, treatment $T_i \in \{0,1\}$, and observed outcome $Y_i = T_i Y_i(1) + (1-T_i)Y_i(0)$, the ITE is $D_i = Y_i(1) - Y_i(0)$. The goal is to find $\hat{\tau}$ minimizing $\mathbb{E}[(D_i - \hat{\tau}(X_i))^2]$, which is equivalent to minimizing the CATE estimation error $\mathbb{E}[(\tau(X_i) - \hat{\tau}(X_i))^2]$.

\subsection{Graph-Based ITE Estimation}
Let $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ with adjacency $\mathcal{A}$. Under network consistency, $Y_i = Y_i(T_i, \mathcal{T}_i)$ where $\mathcal{T}_i$ captures multi-hop neighbor treatment influence. Under standard assumptions (network unconfoundedness, overlap, neighborhood interference), the objective becomes:
\begin{equation}
    \arg\min_{\hat{\tau}} \mathbb{E}[(\tau - \hat{\tau})^2 \mid X=X_i, X_\mathcal{N}=X_\mathcal{N}^i].
\end{equation}

\subsection{X-Learner}
The X-learner estimates $\mu_t = \mathbb{E}[Y \mid T=t, X, X_\mathcal{N}]$ for $t \in \{0,1\}$, computes cross-residuals $\mathrm{D}_1 = Y(1) - \mu_0$ and $\mathrm{D}_0 = \mu_1 - Y(0)$, fits regressors $g_0, g_1$ on these residuals, and combines them: $\hat{\tau} = e \cdot g_0 + (1-e) \cdot g_1$, where $e$ is the propensity score.

%% ====================================================================
\section{Method}
%% ====================================================================

We present two plug-in modules. Crucially, neither modifies the internal architecture of any base model—they operate at the \emph{input} and \emph{loss} levels respectively.

\subsection{Graph Positional Encoding (GPE)}
\label{sec:gpe}

Standard GNNs aggregate neighbor features but lack explicit awareness of a node's global position in the graph. Nodes in the same community may have similar local neighborhoods but different global roles. GPE addresses this by fusing raw node features $h \in \mathbb{R}^{n \times d_h}$ with node2vec~\cite{grover2016node2vec} positional embeddings $p \in \mathbb{R}^{n \times d_p}$ via multi-head self-attention:
\begin{align}
    Q &= hW_Q, \quad K = pW_K, \quad V = pW_V, \\
    \text{head}_i &= \text{softmax}(Q_iK_i^T / \sqrt{d_k})V_i, \\
    \text{GPE}(h, p) &= \text{LayerNorm}([h; \text{MHA}(Q,K,V)] + f_{\text{res}}([h; p])),
\end{align}
where MHA denotes multi-head attention, $[\cdot;\cdot]$ is concatenation, and $f_{\text{res}}$ is a learned residual projection. The output replaces $h$ as input to any downstream GNN backbone.

\textbf{Plug-in interface}: For any model $M$ that takes node features $X$ as input, the GPE-enhanced version is simply $M_{\text{GPE}}(X) = M(\text{GPE}(X, p))$.

\subsection{Variance-Weighted Learning}
\label{sec:var}

Under heteroscedasticity, the ITE can be decomposed as $D_i = \tau(X_i) + \varepsilon_i$ where $\text{Var}[\varepsilon_i \mid X_i] = \sigma^2(X_i)$ varies across individuals. Standard MSE loss treats all samples equally, allowing high-variance samples to dominate gradient updates.

We propose inverse-variance weighting: $w_i = 1/\max\{\hat{\sigma}^2(X_i), \delta\}$, with the weighted regression:
\begin{equation}
    \hat{\tau} = \arg\min_{\hat{\tau}} \sum_{i=1}^n w_i (\mathrm{D}_i - \hat{\tau}(X_i))^2.
\end{equation}

\textbf{For continuous outcomes}, $\sigma^2$ is estimated by a dedicated MLP head that predicts $\log \sigma^2$ (exponentiated when needed), with target $\mathrm{D}_t^2 - \hat{\tau}_t^2$.

\textbf{For binary outcomes} (e.g., click/conversion in recommendation), the noise variance is analytically determined by the Bernoulli model: $\sigma^2_t(X_i) = \mu_t(X_i)(1 - \mu_t(X_i))$, requiring no additional parameters.

\textbf{Plug-in interface}: For any model $M$ with an outcome loss $\mathcal{L}$, the variance-weighted version replaces $\mathcal{L}$ with $\mathcal{L}_{\text{var}} = \sum_i w_i \mathcal{L}_i$, where weights are derived from the model's own predictions (detached to prevent gradient contamination).

\begin{proposition}
Under heteroscedasticity, the variance-weighted estimator achieves $\text{Var}(\hat{\beta}_{\text{WLS}}) \preceq \text{Var}(\hat{\beta}_{\text{OLS}})$ in the Löwner order, i.e., it is the Best Linear Unbiased Estimator (BLUE) by the Gauss-Markov theorem~\cite{Aitken_1936}. Full proof is provided in the code repository.
\end{proposition}

\begin{algorithm}[t]
    \caption{Plug-in Training (for any base model $M$)}
    \label{alg:plugin}
    \begin{algorithmic}[1]
        \STATE \textbf{Input:} Graph $\mathcal{G}$, data $\{X_i, T_i, Y_i\}$, base model $M$
        \STATE \textbf{Optional:} GPE module, variance weighting
        \STATE Compute positional embeddings $p$ via node2vec
        \FOR{each epoch}
            \STATE $\tilde{X} \leftarrow \text{GPE}(X, p)$ if GPE enabled, else $\tilde{X} \leftarrow X$
            \STATE $\hat{\mu}_t, \hat{\tau}_t, \hat{e} \leftarrow M(\tilde{X}, \mathcal{A})$
            \STATE Compute outcome loss $\mathcal{L}$ (model-specific)
            \IF{variance weighting enabled}
                \STATE $w_i \leftarrow 1/(\hat{\sigma}^2(X_i) + \delta)$
                \STATE $\mathcal{L} \leftarrow \sum_i w_i \mathcal{L}_i / \sum_i w_i$
            \ENDIF
            \STATE Update $M$ parameters via $\nabla \mathcal{L}$
        \ENDFOR
    \end{algorithmic}
\end{algorithm}

%% ====================================================================
\section{Experiments}
%% ====================================================================

\subsection{Experimental Setup}

\textbf{Graph datasets.} We use five graph datasets with semi-synthetic outcomes: CoraFull (19,793 nodes), DBLP (17,716), PubMed (19,717)—with heteroscedastic noise $\rho \in \{5, 10, 30\}$—and the standard BlogCatalog (5,196) and Flickr (7,575) benchmarks from~\cite{guo2020learning}. Our semi-synthetic data generation extends prior work by incorporating Louvain-detected communities, multi-hop treatment spillover, and per-node heteroscedastic noise.

\textbf{Tabular datasets.} We evaluate on six real-world uplift benchmarks without graph structure: Criteo Uplift (1M rows, binary conversion), Hillstrom (42K, binary visit and continuous spend), RetailHero X5 (200K, binary purchase), Lenta (690K, binary response), and X5 with enriched purchase features.

\textbf{Base models.} We test eight models spanning the full timeline of graph ITE research:
\begin{itemize}
    \item \emph{Meta-learners}: BNN (S-learner)~\cite{johansson2016learning}, TARNet (T-learner)~\cite{shalit2017estimating}, X-learner~\cite{kunzel2019metalearners}
    \item \emph{SOTA graph models}: NetDeconf~\cite{guo2020learning}, GIAL~\cite{chu2021graph}, GNUM~\cite{zhu2023gnum}, GDC~\cite{hu2025gdc}
    \item \emph{Combined}: CAVIN (X-learner + GPE + Var)
\end{itemize}

\textbf{Evaluation.} We report $\sqrt{\epsilon_{\text{PEHE}}}$ (lower is better) and Qini coefficient (higher is better). All graph experiments use 3 seeds; tabular experiments use 5 seeds. Training uses Adam with lr=$10^{-3}$, 200 epochs, 60/20/20 splits.

\subsection{GPE as Universal Plug-in (Graph)}

Table~\ref{tab:gpe_pehe} shows the effect of adding GPE to each base model. We highlight three key observations:

\begin{table}[t]
\caption{GPE plug-in effect on PEHE ($\downarrow$). Values are mean over 3 seeds. Bold = best per row. ``$\Delta$'' shows relative improvement. Results on CoraFull/DBLP/PubMed at $\rho=10$.}
\label{tab:gpe_pehe}
\small
\begin{tabular}{lccr}
\toprule
Model & Vanilla & +GPE & $\Delta$\% \\
\midrule
TARNet & 3.00 / 2.52 / 3.40 & \textbf{2.64 / 2.27 / 3.23} & $-$12 / $-$10 / $-$5 \\
GDC & 3.91 / 3.19 / 3.06 & \textbf{3.62 / 2.70 / 2.82} & $-$7 / $-$15 / $-$8 \\
NetDeconf & 7.65 / 3.92 / 3.33 & \textbf{4.64 / 3.72 / 3.30} & $-$39 / $-$5 / $-$1 \\
GIAL & 6.18 / 4.11 / 3.37 & \textbf{6.27 / 3.69 / 3.35} & — / $-$10 / $-$1 \\
GNUM & 6.09 / 3.89 / 4.10 & \textbf{4.98 / 3.82 / 4.05} & $-$18 / $-$2 / $-$1 \\
X-learner & 3.21 / 2.31 / 3.09 & \textbf{2.62 / 2.25 / 3.17} & $-$18 / $-$3 / — \\
BNN & 5.47 / 5.02 / 5.45 & 5.53 / 5.03 / 5.45 & — \\
\bottomrule
\end{tabular}
\end{table}

\textbf{(1) GPE improves 7 of 8 base models.} Only BNN (S-learner) fails to benefit, because its shared $\mu(x,t)$ head concatenates the treatment indicator with features, preventing positional information from differentially conditioning treatment vs.\ control representations.

\textbf{(2) TARNet+GPE is the most consistent winner.} Across all 9 (dataset, $\rho$) cells on CoraFull/DBLP/PubMed plus 2 on BlogCatalog/Flickr, TARNet+GPE achieves a perfect 11/11 PEHE improvement.

\textbf{(3) GPE improves the current SOTA.} GDC+GPE reduces PEHE by 15\% on DBLP ($\rho=5$: 3.12$\to$2.60) and by 8\% on PubMed ($\rho=10$: 3.06$\to$2.82).

\begin{table}[t]
\caption{GPE plug-in win rate (PEHE improvement count out of 9 dataset$\times\rho$ cells on CoraFull/DBLP/PubMed, and 2 cells on BlogCatalog/Flickr).}
\label{tab:gpe_winrate}
\small
\begin{tabular}{lcc}
\toprule
Base Model $\to$ +GPE & Our Datasets (9) & Blog/Flickr (2) \\
\midrule
TARNet & \textbf{9/9} & \textbf{2/2} \\
NetDeconf & 7/9 & 1/2 \\
GDC (SOTA) & 7/9 & 1/2 \\
GIAL & 7/9 & 0/2 \\
X-learner & 7/9 & 1/2 \\
GNUM & 6/9 & 2/2 \\
CAVIN & 6/9 & 1/2 \\
BNN & 1/9 & 0/2 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Variance Weighting as Plug-in (Graph)}

Table~\ref{tab:var_graph} shows variance weighting results on graph data. The module is most effective at high noise ($\rho=30$) where it rescues GPE instability:

\begin{table}[t]
\caption{GPE + Variance weighting synergy on GDC at $\rho=30$. PEHE ($\downarrow$), mean over 3 seeds.}
\label{tab:var_graph}
\small
\begin{tabular}{lcccr}
\toprule
Dataset & GDC & +GPE & +Var & +GPE+Var \\
\midrule
CoraFull & 4.93 & 5.90 & 4.32 & \textbf{4.21} \\
DBLP & 4.26 & 3.91 & 4.30 & \textbf{3.57} \\
PubMed & 4.31 & 4.80 & 3.79 & \textbf{2.99} \\
\bottomrule
\end{tabular}
\end{table}

At $\rho=30$, GPE alone \emph{hurts} GDC on CoraFull (4.93$\to$5.90) and PubMed (4.31$\to$4.80)—the positional attention overfits to noisy labels. Adding variance weighting rescues this: GDC+GPE+Var achieves 2.99 on PubMed, a 31\% reduction from vanilla.

\subsection{Real-World Tabular Validation}

To validate variance weighting without graph structure, we test on six tabular benchmarks using S-, T-, and X-learner with and without Bernoulli variance weighting:

\begin{table}[t]
\caption{Tabular variance weighting (Qini $\uparrow$, 5 seeds). Bold = best per dataset.}
\label{tab:tabular}
\small
\begin{tabular}{lcccc}
\toprule
Dataset & Outcome & S & S+Var & $\Delta$ \\
\midrule
Hillstrom & spend (cont.) & 0.029 & \textbf{0.117} & $+$4$\times$ \\
X5 RetailHero & binary & 0.013 & 0.013 & — \\
Criteo & binary & \textbf{0.167} & 0.151 & — \\
Hillstrom & binary & $-$0.001 & $-$0.005 & — \\
\bottomrule
\end{tabular}
\end{table}

The standout result is Hillstrom (continuous spend): S+Var achieves a 4$\times$ Qini improvement (0.029$\to$0.117). This confirms that variance weighting is most effective when the outcome exhibits genuine heteroscedasticity—continuous outcomes with varying noise levels across individuals. For binary outcomes where $\text{Var}[Y] = \mu(1-\mu)$ is bounded, the improvement is more modest but variance weighting consistently reduces cross-seed standard deviation (a stability benefit relevant for production deployment).

\subsection{Ablation: Component Interactions}

\begin{table}[t]
\caption{Full ablation on DBLP $\rho=30$ (PEHE $\downarrow$). Each cell: +GPE adds GPE, +Var adds variance weighting.}
\label{tab:ablation}
\small
\begin{tabular}{lcccc}
\toprule
Model & Vanilla & +GPE & +Var & +GPE+Var \\
\midrule
TARNet & 6.33 & 5.26 & 6.51 & \textbf{4.27} \\
X-learner & 4.66 & 4.61 & \textbf{3.84} & 4.14 \\
GDC & 4.26 & 3.91 & 4.30 & \textbf{3.57} \\
NetDeconf & 4.51 & \textbf{4.35} & 5.04 & 4.93 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:ablation} reveals the interaction patterns: (1) For TARNet and GDC, the \emph{combination} of GPE and Var outperforms either alone—the two modules address complementary failure modes (structural bias vs.\ outcome noise). (2) For X-learner, Var alone achieves the best result—the X-learner's cross-fitting already captures some structural information, making GPE less critical. (3) For NetDeconf, GPE alone works best—adding variance weighting to this Wasserstein-balanced model disrupts its representation learning.

%% ====================================================================
\section{Conclusion}
%% ====================================================================

We have shown that Graph Positional Encoding and Variance-Weighted Learning are effective, orthogonal, and architecture-agnostic plug-in modules for graph-based uplift modeling. GPE improves 7 of 8 tested base models including the 2025 SOTA, while variance weighting provides complementary gains under heteroscedastic noise. Together, they reduce GDC's PEHE by up to 31\%. On real-world tabular data, variance weighting achieves a 4$\times$ Qini improvement on continuous outcomes.

Rather than proposing another monolithic model, our work demonstrates a \emph{modular enhancement paradigm}: researchers can apply GPE and variance weighting to their preferred base model and expect consistent improvements. Code and full experimental results are available at \url{https://anonymous.4open.science/r/CAVIN}.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
